SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-03-29 paper/paper.tex
final paper updates.

\begin{abstract}
QuickCheck, developed by Claessen and Hughes, is a powerful library for
automatic test-case generation.  Because QuickCheck performs random testing,
some of the counterexamples discovered are very large.  QuickCheck provides an
interface for the user to write \emph{shrink} functions to attempt to reduce the
size of counterexamples discovered.  Hand-written implementations of shrink can
be complex, inefficient, and consist of significant boilerplate code.  We
introduce \emph{SmartCheck}, an automatic efficient counterexample reduction
algorithm.  SmartCheck reduces algebraic data using generic search heuristics to
efficiently find smaller counterexamples independent of the program or property
being tested.  In addition to shrinking, SmartCheck generalizes counterexamples
by returning a \emph{formula} that abstracts over individual counterexamples and
make it easier for the programmer to uncover multiple counterexamples at once.
SmartCheck has been implemented for Haskell and is freely available.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Originally designed for Haskell,
QuickCheck has been ported to other languages and is a now a widely-used testing
tool.  Because QuickCheck generates random values for testing, counterexamples
it finds may be substantially larger than a minimal counterexample.  In their
original QuickCheck paper~\cite{qc}, the authors report the following user
experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a]
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values, according
to some measure, from a given counterexample.  These new values are then tested
to attempt to find a smaller counterexample.  Today, \ttp{smaller} is called
\ttp{shrink}.

@@

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is generalization.  For
example, the following is a relatively small counterexample returned when
QuickChecking a property in (a buggy version of) Xmonad, a popular X11 window
manager written in Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Functional programmers should be familiar with having to debug
a ``wall of text'' when debugging data structures.  What if instead a
counterexample like the following were returned, stating that for any well-typed
values $x_0$, $x_1$, $x_2$, and $x_3$, a counterexample is returned:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
The formula quantifies away all the irrelevant portions of the data structure
with respect to the property, so that the user can focus on the heart of the
problem in the counterexample.

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed \emph{SmartCheck}.  SmartCheck takes a counterexample produced by some
oracle---we use QuickCheck---and generically minimizes and generalizes the
counterexample.

As demonstrated by a motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  More importantly, the
time required to shrink scales linearly with the size of the
counterexample---other common shrink implementations often scale in polynomial
time or worse.  (Shrinking time is important in real-world uses; see
Section~\ref{sec:experiments} for examples.)  And the counterexamples returned
are two to five times smaller.

@@

SmartCheck implements three novel approaches to automatically generalize
counterexamples, which are described in Section~\ref{sec:generalization}.  The
first algorithm universally quantifies sub-values that always fail in tests, as
briefly described earlier in the introduction.  The second algorithm
existentially quantifies sub-values for types in which every possible variant
fails the property.  For example, finding counterexamples (\ttp{Left 2}) and
(\ttp{Right True}) for the type
%
\begin{code}
\ttp{Either Int Bool}
\end{code}
%
\noindent
means there exists a counterexample regardless of the variant chosen.

The third algorithm automatically strengthens properties by omitting ``similar''
counterexamples to the ones previously observed.  The algorithm is motivated by
noting that there are often multiple ways in which a property may fail; for
example, a property that parsing a pretty-printed abstract syntax tree (AST)
results in the original AST may fail due to multiple bugs.  During testing, it
is useful to discover counterexamples arising from each bug in one go.  In
practice, the problem is solved by discovering a counterexample \ttp{cex},
abstracting it, and then adding a new precondition to the property that
informally says ``omit counterexamples of form \ttp{cex}.''  Adding
preconditions manually is laborious and may cause the programmer to make
premature fixes to the program, if she believes she has isolated the error.

@@

In this section, we motivate in more detail the challenges in shrinking
counterexamples (we focus on shrinking and only touch on generalization here).
Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing four lists of
wrapped \ttp{Int16}s and a single wrapped \ttp{Word8}.  For our purposes, it is
not important what \ttp{T} models, but for concreteness, suppose it models the
values in four linked lists in an operating system, together with an 8-bit
status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (the \ttp{==>}) operator in the figure
implements implication from the QuickCheck library).

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable, there are two problems with it.
First, the result of (\ttp{shrink t}) is null if any list contained in \ttp{t}
is null, so \ttp{t} will not be shrunk.  More troublesome is that with
QuickCheck's default \ttp{Arbitrary} instances, the length of potential
counterexamples returned by \ttp{shrink} can easily exceed $10^{10}$, which is
an intractable number of tests for QuickCheck to analyze.  The reason for the
blowup is that shrinking a list \ttp{[a]} produces a list \ttp{[[a]]}, the
length of which is significantly longer than the original list.  Then, we take
the cross-product of the generated lists.  For the example above, with some
counterexamples, the shrinking stage may appear to execute for hours, consuming
ever more memory, without returning a result.

@@

Still, using this definition, from some initial counterexamples, shrinking may
take over 15 seconds and result in a final counterexample containing nearly 60
\ttp{N} constructors.\footnote{All results reported in the paper are with a
  GHC-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7 processor.}

SmallCheck is another testing framework for Haskell for which shrinking is
irrelevant.  SmallCheck is guaranteed to return a smallest counterexample, if
one exists~\cite{sc}.  SmallCheck does this by enumerating all possible inputs,
ordered from smallest to largest, up to some user-defined bound.  While
SmallCheck is effective for testing many programs and properties (in accordance
with the \emph{small scope hypothesis}~\cite{jackson}), counterexamples to even
relatively simple properties may be infeasible to discover due to input-space
explosion.

@@

To motivate the benefits of SmartCheck reduction algorithm, consider
Table~\ref{table:results} and the corresponding graphs in
Figure~\ref{fig:graphs}, giving the results for using QuickCheck and SmartCheck
on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck, since as
noted, it is not even feasible for the example).  We graph the final
counterexample size and execution time.  Out of 2000 runs, we show the number of
runs resulting in a shrunk counterexample of a given size (by counting \ttp{N}
constructors), and we show the number of runs executing within a given time
limit (in seconds).  Note that we provide two execution-time graphs, since time
scales differ dramatically (we use a logrithmic $y$ axis in one graph).  In the
table, we show the mean, median, median, and the results at the 95th percentile
values.  (The distributions produced by these values are not necessarily
Gaussian, so we omit the standard deviation.)

@@

More dramatic, however, are the differences in execution time.  The execution
times for SmartCheck include the initial counterexample generation time required
by QuickCheck (QC none); thus the time required by SmartCheck to shrink a
counterexample is on average just two hundredths of a second (0.10 - 0.08).  As
can be seen in the graphs, SmartCheck's execution time scales linearly (like
QuickCheck without shrinking).  Truncation and tuple shrinking produce very
long tails; the longest execution time recorded for truncation shrinking and
tuple shrinking are 88 seconds and 110 seconds, respectively.  SmartCheck is
largely insensitive to the size of the original counterexample provided.

The results for SmartCheck are fairly generalizable to other types and
properties since SmartCheck's reduction algorithm has no dependencies on the
specifics of either; however, user-defined shrink methods may vary significantly
based on the property and counterexample type.

Finally, note that after shrinking the counterexample, SmartCheck will also
generalize it.  Note that for any \ttp{Word8} value in the first argument of
\ttp{T}, the counterexample still holds.  For example, SmartCheck might return a
counterexample of the form
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}

@@

\section{Shrinking Data}\label{sec:shrinking}
In this section, we describe how to efficiently and generically shrink algebraic data
values.  Before we begin, recall the basic idea behind the \ttp{shrink} method
of the \ttp{Arbitrary} class: generate a list of values, each of which is
smaller than the current counterexample.  Each of the new values generated
typically bears no relationship to the original counterexample other than being
smaller.

Consider the problem of finding a small counterexample as a graph search
problem.  The set of vertexes are all possible well-typed values.  There is an
edge from vertex $v$ to $v'$ if $v'$ is smaller than $v$ by some measure.  The
goal is to traverse the graph, starting from a counterexample, to find a vertex
from the set of smallest counterexamples in the graph.

The approach taken by typical \ttp{shrink} implementations is the most naive
graph search algorithm: it simply generates a random set of potential vertexes
that are strictly smaller according to the given measure to check the property
against.  However, it does not exploit that the original counterexample
\ttp{cex} is already in a neighborhood of other likely counterexamples.  So
rather than choose random vertexes, it may be more efficient to test other
values that are closely related to \ttp{cex}.  By restricting the search to
\ttp{cex}'s neighborhood, a smaller counterexample might be missed, and only a
local minimum is returned.  However, with a large set of values, finding a
global minimum by random sampling vertexes in the graph is possible but too
inefficient.

@@

\subsection{Reduction Algorithm Details}
Having described the reduction algorithm, there are two important details about
its design we describe below.

@@

Consequently, we distinguish a class of types referred to as \emph{opaque
  types}.  If the reduction algorithm encounters a opaque type, it is ignored.
By default, this includes types without constructors (e.g., \ttp{Int}, \ttp{Word},
\ttp{Float}, \ttp{Bool}, and \ttp{Char}).  We also treat recursive container
types like lists specially: lists of opaque types are considered to be opaque
types themselves.  This prevents the algorithm from analyzing, for example,
strings.  A list of non-opaque-types, however, is shrunk by the algorithm.

By providing custom instances, the user can declare any substructure in a data
type to be a opaque type.  Doing so effectively treats values from that type as
``black boxes'' for SmartCheck.  Doing so can help make SmartCheck more
efficient if the user knows that some portion of the structure cannot be shrunk
or is irrelevant to the property.

@@

\subsection{Reduction Algorithm Optimizations}
The reduction algorithm description above omits some details and optimizations
we describe here.

@@

eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1

prop_div e = divSubTerms e ==> eval e /= Nothing
\end{code}
%
Testing \ttp{prop\_div}, we might have a counterexample like the following:
%
\medskip%
\begin{sloppypar}
\small
\noindent%
\ttp{Add (Div (C 5) (C (-12))) (Add (Add (C 2) (C 4)) (Add (C 7) (Div (Add (C 7)
  (C 3)) (Add (C (-5)) (C 5)))))}
\end{sloppypar}
\medskip%
\noindent
In the counterexample, the culprit is a buried sub-value:
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
Thus, when attempting to shrink an \ttp{Exp} value, it pays to test whether a
sub-value itself fails the property.

@@

An extrapolation algorithm performs universal sub-value generalization.  The
basic idea is as follows: for a counterexample \ttp{cex} and a property
\ttp{prop}, a breadth-first search over the sub-values of the \ttp{cex} is
performed.  For each sub-value, the algorithm generates new sub-values and replaces
them in \ttp{cex} to create a list of new potential counterexamples.  If enough
new counterexamples non-trivially fail \ttp{prop} (i.e., fail the property's
consequent but satisfy its antecendent), then we extrapolate, claiming that for
\emph{any} new value replacing that sub-value in \ttp{cex}, the property will
fail.

@@

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if enough
new values pass the property's precondition and fail the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is roughly the dual of \ttp{pass} in the reduction algorithm: (\ttp{fail
  prop cex}) returns (\ttp{Just True}) if \ttp{cex} passes \ttp{prop}'s
precondition but fails the property; (\ttp{Just False}) if \ttp{cex}
non-trivially satisfies \ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails
\ttp{prop}'s precondition.  The flag \ttp{scMinForall} is the minimum number
of \ttp{Just False} results required from \ttp{fail} to extrapolate from failed
tests to a universal claim.

@@

The function \ttp{constrFail} returns true if we replace the sub-value at index
\ttp{idx} in counterexample \ttp{cex} with every possible variant given the type
and construct a counterexample to the property.  There are four guards to the
recursive function \ttp{constrFail'}: the first guard holds if the list of
constructors tagging variants in which a counterexample is found is equal in
size to the list of all possible constructors for the type.  The second guard
tests whether the set of test values is null; if so (and if the first guard
fails), then we have exhausted test values before finding all possible failing
variants.  Third, for a specific sub-value \ttp{v}, we test whether it fails the
property.  If so, we add its constructor to the list of constructors.
Otherwise, we simply recurse.  Note in the definition of \ttp{prop'}, we add an
additional precondition that the current constructor is not an element of
constructors already seen.  Thus, (\ttp{go v}) returns (\ttp{Just True}) if
(\ttp{replace cex idx v}) passes this precondition (and any other preconditions
of \ttp{prop}), but fails the property.

>>>>

\begin{abstract}
QuickCheck, developed by Claessen and Hughes, is a powerful library for automatic test-case generation.  Because QuickCheck performs random testing, some of the counterexamples discovered are very large.  QuickCheck provides an interface for the user to write shrink functions to attempt to reduce the size of counterexamples discovered.  Hand-written implementations of shrink can be complex, inefficient, and consist of significant boilerplate code.  Furthermore, shrinking is only one aspect in debugging: counterexample generalization is the process of extrapolating from individual counterexamples to a class of counterexamples, often requiring a flash of insight from the programmer.  To improve counterexample reduction and generalization, we introduce \emph{SmartCheck}.  SmartCheck is a debugging tool that reduces algebraic data using generic search heuristics to efficiently find smaller counterexamples independent of the program or property being tested.  In addition to shrinking, SmartCheck also automatically generalizes counterexamples to formulas representing them.  SmartCheck has been implemented for Haskell and is freely available.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc,monadic}.  Originally designed for Haskell,
QuickCheck has been ported to other languages and is a now a widely-used testing
tool.  Because QuickCheck generates random values for testing, counterexamples
it finds may be substantially larger than a minimal counterexample.  In their
original QuickCheck paper~\cite{qc}, the authors report the following user
experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a]
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values, according
to some measure, from a given counterexample.  These new values are then tested
to attempt to find a smaller counterexample.  Today, \ttp{smaller} is called
\ttp{shrink}.

@@

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, the following is a
relatively small counterexample returned when QuickChecking a property in (a
buggy version of) Xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Functional programmers should be familiar with having to debug
a ``wall of text'' when debugging data structures.  What if instead a
formula like the following were returned, stating that for any well-typed
values $x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

%
\begin{samepage}
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
\end{samepage}
%
\noindent
The formula quantifies away all the irrelevant portions of the data structure
with respect to the property, so that the user can focus on the heart of the
problem in a class of counterexamples.

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed \emph{SmartCheck}.  SmartCheck takes a counterexample produced by some
oracle (we use QuickCheck) and generically minimizes and generalizes the
counterexample.

As demonstrated by a motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  As well, the time
required to shrink scales linearly with the size of the counterexample---other
common shrink implementations often scale in polynomial time or worse.
(Shrinking time is important in real-world uses; see
Section~\ref{sec:experiments} for examples.)  And the counterexamples returned
are two to five times smaller.

@@

SmartCheck implements three novel approaches to automatically generalize
counterexamples, which are described in Section~\ref{sec:generalization}.  The
first algorithm universally quantifies sub-values that always fail in tests, as
briefly described earlier in the introduction.  The second algorithm
existentially quantifies sub-values for types in which every possible variant
fails the property.  For example, finding counterexamples (\ttp{Left 2}) and
(\ttp{Right True}) for the type
%
\begin{code}
\ttp{Either Int Bool}
\end{code}
%
\noindent
means there exists a counterexample regardless of the variant chosen.
Existential generalization is important for large sum types, as found in
abstract syntax tree (AST) definitions, for example.

The third algorithm automatically strengthens properties by omitting ``similar''
counterexamples to the ones previously observed.  The algorithm is motivated by
noting that there are often multiple ways in which a property may fail; for
example, a property that parsing a pretty-printed AST results in the original
AST may fail due to multiple bugs.  During testing, it is useful to discover
counterexamples arising from each bug in one go.  In practice, the problem is
solved by discovering a counterexample \ttp{cex}, abstracting it, and then
adding a new precondition to the property that informally says ``omit
counterexamples of form \ttp{cex}.''  Adding preconditions manually is laborious
and may cause the programmer to make premature fixes to the program, if she
believes she has isolated the error before she actually does.

@@

In this section, we motivate in more detail the challenges in shrinking
counterexamples (we focus on shrinking and only touch on generalization here).
Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.  We use Haskell \ttp{Prelude} functions, \ttp{mapMaybe} from
  \ttp{Data.Maybe}, and \ttp{liftM} and \ttp{replicateM} from
  \ttp{Control.Monad}.}  Data type \ttp{T} is a product type containing four
lists of wrapped \ttp{Int16}s and a single wrapped \ttp{Word8}.  For our
purposes, it is not important what \ttp{T} models, but for concreteness, suppose
it models the values in four linked lists in an operating system, together with
an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (the \ttp{==>}) operator in the figure
is implication from the QuickCheck library).

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable, there are two problems with it.
First, the result of (\ttp{shrink t}) is null if any list contained in \ttp{t}
is null, in which case \ttp{t} will not be shrunk.  More troublesome is that with
QuickCheck's default \ttp{Arbitrary} instances, the length of potential
counterexamples returned by \ttp{shrink} can easily exceed $10^{10}$, which is
an intractable number of tests for QuickCheck to analyze.  The reason for the
blowup is that shrinking a list \ttp{[a]} produces a list \ttp{[[a]]}, the
length of which is significantly longer than the original list.  Then, we take
the cross-product of the generated lists.  For the example above, with some
counterexamples, the shrinking stage may appear to execute for hours, consuming
ever more memory, without returning a result.

@@

Still, using this definition, from some initial counterexamples, shrinking may
take over one minute and result in a final counterexample containing just under
100 \ttp{N} constructors.\footnote{All results reported in the paper are with a
  GHC-7.6.1-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7
  processor.}

SmallCheck is another testing framework for Haskell for which shrinking is
irrelevant: SmallCheck is guaranteed to return a smallest counterexample, if
one exists~\cite{sc}.  SmallCheck does this by enumerating all possible inputs,
ordered from smallest to largest, up to some user-defined bound.  While
SmallCheck is effective for testing many programs and properties (in accordance
with the \emph{small scope hypothesis}~\cite{jackson}), counterexamples to even
relatively simple properties may be infeasible to discover due to input-space
explosion.

@@

To motivate the benefits of SmartCheck reduction algorithm, consider
Table~\ref{table:results} and the corresponding graphs in
Figure~\ref{fig:graphs}, giving the results for using QuickCheck and SmartCheck
on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck, since as
noted, it is not even feasible for the example with default instances).  We
graph the final counterexample size and execution time.  Out of 2000 runs, we
show the number of runs resulting in a shrunk counterexample of a given size (by
counting \ttp{N} constructors), and we show the number of runs executing within
a given time limit (in seconds).  Note that we provide two execution-time
graphs, since time scales differ dramatically (we use a logrithmic scale for the
$y$ axis in one graph and a linear scale in the other).  In the table, we show
the mean, median, median, and the results at the 95th percentile values.  (The
distributions produced by these values are not necessarily Gaussian, so we omit
the standard deviation.)

@@

More dramatic, however, are the differences in execution time.  The execution
times for SmartCheck include the initial counterexample generation time required
by QuickCheck (QC none); thus the time required by SmartCheck to shrink a
counterexample is on average just two hundredths of a second (0.10 - 0.08).  As
can be seen in the graphs, SmartCheck's execution time scales linearly (like
QuickCheck without shrinking).  Truncation and tuple shrinking produce very long
tails; the longest execution time recorded for truncation shrinking and tuple
shrinking are 88 seconds and 110 seconds, respectively, although these outliers
can vary wildly.  SmartCheck is largely insensitive to the size of the original
counterexample provided.

In regards to generalization, after shrinking the counterexample, SmartCheck
will also generalize it.  Note that for any \ttp{Word8} value in the first
argument of \ttp{T}, the counterexample still holds.  For example, SmartCheck
might return a counterexample of the form
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}

Finally, these performance results for SmartCheck are fairly generalizable to
other types and properties, since the algorithm is not dependent on the specific
program or property being checked.  However, the results are particularly
pronounced for counterexamples that are particularly ``deep'' or ``broad'' in
terms of their tree structure.  For programs and properties for which small
counterexamples are easily found, QuickCheck's shrinking is perfectly
acceptable.  Furthermore, user-defined shrink methods may vary significantly
based on the property and counterexample type, so it is not always clear which
implementation to compare against.

@@

\section{Shrinking Data}\label{sec:shrinking}
In this section, we describe how to efficiently and generically shrink algebraic
data values.  Before we begin, recall the basic idea behind the \ttp{shrink}
method of the \ttp{Arbitrary} class: generate a list of values, each of which is
smaller than the current counterexample.  Each of the new values generated
may not bear any relationship to the original counterexample other than being
smaller.  SmartCheck pursues an approach that searches for smaller but
structurally similar counterexamples, as we make precise below.

We describe the algorithm in Section~\ref{sec:reduct} and then describe
algorithmic details in Section~\ref{sec:details}.  Some optimizations to the
reduction algorithm are described in Section~\ref{sec:optimizations}.

%% Consider the problem of finding a small counterexample as a graph search
%% problem.  The set of vertexes are all possible well-typed values.  Let $<$ be
%% a partial ordering on values relating them by some size measure.  Let there
%% be an edge from vertex $v$ to $v'$ if $v < v'$ and there exists no $v''$ such
%% that $v < v'' < v'$.  The goal is to traverse the graph, starting from a
%% counterexample, to find a vertex from the set of smallest counterexamples in
%% the graph.

%% The approach taken by typical \ttp{shrink} implementations is the most naive
%% graph search algorithm: it simply generates a random set of potential vertexes
%% that are strictly smaller according to the given measure to check the property
%% against.  However, it does not exploit that the original counterexample
%% \ttp{cex} is already in a neighborhood of other likely counterexamples.  So
%% rather than choose random vertexes, it may be more efficient to test other
%% values that are closely related to \ttp{cex}.  By restricting the search to
%% \ttp{cex}'s neighborhood, a smaller counterexample might be missed, and only a
%% local minimum is returned.  However, with a large set of values, finding a
%% global minimum by random sampling vertexes in the graph is possible but too
%% inefficient.

@@

\subsection{Reduction Algorithm Details}\label{sec:details}
Having described the reduction algorithm, there are two important details about
its design we describe below.

@@

Consequently, we distinguish a class of types referred to as \emph{opaque
  types}.  If the reduction algorithm encounters a opaque type, it is ignored.
By default, this includes ``primitive'' types (e.g., \ttp{Int}, \ttp{Word},
\ttp{Float}, \ttp{Bool}, and \ttp{Char}).  We also treat recursive container
types like lists specially: lists of opaque types are considered to be opaque
types themselves.  This prevents the algorithm from analyzing, for example,
strings.  A list of non-opaque-types, however, is shrunk by the algorithm.

By providing custom instances, the user can declare any substructure in a data
type to be a opaque type.  Doing so effectively treats values from that type as
``black boxes'' for SmartCheck, making SmartCheck more efficient if the user
knows that some portion of the structure cannot be shrunk or is irrelevant to
the property.

@@

\subsection{Reduction Algorithm Optimizations}\label{sec:optimizations}
The reduction algorithm description above omits some details and optimizations
we describe here.

@@

eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
odivSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1

prop_div e = divSubTerms e ==> eval e /= Nothing
\end{code}
%
Testing \ttp{prop\_div}, we might have a counterexample like the following:
%
\medskip%
\begin{sloppypar}
\small
\noindent%
\ttp{Add (Div (C 5) (C (-12))) (Add (Add (C 2) (C 4)) (Add (C 7) (Div (Add (C 7)
  (C 3)) (Add (C (-5)) (C 5)))))}
\end{sloppypar}
\medskip%
%
\noindent
The cause is that \ttp{divSubTerms} fails to check whether the
divisor evaluates to zero.  In the counterexample, the culprit is a buried
sub-value:
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Thus, when attempting to shrink an \ttp{Exp} value, it pays to test whether a
sub-value itself fails the property.

@@

An extrapolation algorithm performs universal sub-value generalization.  The
basic idea is as follows: for a counterexample \ttp{cex} and a property
\ttp{prop}, a breadth-first search over the sub-values of the \ttp{cex} is
performed.  For each sub-value, the algorithm generates new sub-values and
replaces them in \ttp{cex} to create a list of new potential counterexamples.
If no new value satisfies the property, then we extrapolate, claiming that for
\emph{any} new value replacing that sub-value in \ttp{cex}, the property will
fail.

@@

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if no new
values satisfy the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is roughly the dual of \ttp{pass} in the reduction algorithm: (\ttp{fail
  prop cex}) returns (\ttp{Just True}) if \ttp{cex} passes \ttp{prop}'s
precondition but fails the property; (\ttp{Just False}) if \ttp{cex}
non-trivially satisfies \ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails
\ttp{prop}'s precondition.  The flag \ttp{scMinForall} is the minimum number
of \ttp{Just False} results required from \ttp{fail} to extrapolate from failed
tests to a universal claim.

@@

The function \ttp{constrFail} returns true if we replace the sub-value at index
\ttp{idx} in counterexample \ttp{cex} with every possible variant given the type
and construct a counterexample to the property.  There are four guards to the
recursive function \ttp{constrFail'}: the first guard holds if the list of
constructors tagging variants in which a counterexample is found is equal in
size to the list of all possible constructors for the type.  The second guard
tests whether the set of test values is null; if so (and if the first guard
fails), then we have exhausted test values before finding all possible failing
variants.  Third, for a specific sub-value \ttp{v}, we test whether it fails the
property.  If so, we add its constructor to the list of constructors.
Otherwise, we simply recurse.  Note in the definition of \ttp{prop'}, we add an
additional precondition that the current constructor is not an element of
constructors already seen.  Thus, (\ttp{go v}) returns
%
\begin{code}
\ttp{Just True}
\end{code}
%
\noindent
if
%
\begin{code}
\ttp{replace cex idx v}
\end{code}
%
passes this precondition (and any other preconditions of \ttp{prop}), but fails
the property.
